Ages of Stars and Planets in the Kepler Field Over the First Four Billion Years

### 2024-08-20 papers/paper/ms.tex
[WIP] update ms

Finally, since our scope is focused on rotation-based ages, we
restricted our attention to stars with reported $P_{\rm
rot}$$<$45\,days.  The slowest-rotating FGK stars in the open clusters
used to calibrate our gyrochronology model have $P_{\rm
rot}$$\approx$35\,days.  Figure~\ref{fig:prot_vs_teff} shows a subset
of the resulting \nuniqstarsantosrot\ Kepler stars with rotation
periods compiled from \citetalias{Santos_2019},
\citetalias{Santos_2021}, and our extended KOI list.\added{  The gray
lines in this figure are the polynomials described by \citet{Bouma_2023}
that were
fitted to the slow rotation sequences of the Pleiades \citep{Rebull_2016},
Praesepe \citep{Rampalli_2021}, NGC-6811 \citep{Curtis_2019_ngc6811}, NGC-6819 \citep{Meibom_2015},
Ruprecht-147 \citep{Curtis_2020}, and M67 \citep{Barnes2016,Dungee_2022,Gruner_2023}.}

@@

\subsection{Stellar Quality Flags}
\label{subsec:flags}
We calculated gyrochrone ages for all \nuniqstarsantosrotteffcut\
stars with reported rotation periods that had effective temperatures
of 3800--6200\,K.  To identify stars for which we suspect these ages
may not be valid, we then built a set of quality flags which we
condensed into a single binary number: $Q_{\rm star}$.  When and how
this bitmask should be applied depends on the question being asked.
If the goal is to construct a false-positive free sample, all the
quality flags could be applied.  If the goal is to construct a
complete sample, then consider the examples of Kepler~1627Ab
($\approx$40\,Myr) and Kepler~51c ($\approx$625\,Myr).  The former has
a high RUWE due to a resolved binary companion \citep{Bouma_2022a};
the latter is on a grazing orbit \citep{2014ApJ...783...53M}.  We
leave selection for or against such cases as a decision to the user.
For our own analysis, we assume that a star is suitable for
gyrochronology if none of bits zero through nine (inclusive) are
raised.  For analyses that require all stellar rotation periods to
come from the same detection pipeline, we further require bit 11 to be
zero.

@@

%The rotation period
%distribution of sources excluded in this manner includes a relatively
%larger fraction of ``rapid rotators'' than the main sequence sources,
%likely due to a combination of astrophysical and systematic effects.
%See for instance Figure~8a of \citet{2022AJ....164..137K} or
%Figures~12 and 13 of \citet{2023ApJS..268....4F} for examples of how
%the period distribution of this population differs from stars near the
%main sequence.

{\it Candidate pulsators and close-in binaries (bit
10)}---\citeauthor{Santos_2021} included in a flag for candidate
``classical pulsators'' (e.g.\ Cepheids) and close-in binaries.
Visual inspection shows that although this flag selects many bona fide
objects in these classes, it also selects most known young planets
with $P_{\rm rot}$$\lesssim$5\,day.  These objects are
neither classical pulsators nor close binaries.  While we propagated
this flag into our set of quality flags, we therefore generally ignore
it.

{\it Not in the homogeneous stellar rotation sample (bit 11)}--- This
flag was set for the KOIs with rotation periods drawn from a source
other than the \citeauthor{Santos_2019} pipeline.

@@

	\vspace{-0.35cm}
    \subfloat{
        \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000_preciseagesonly.pdf}
    }
  \end{center}
  \vspace{-0.6cm}
  \caption{{\bf Kepler's demographic cliff,} visible in the
  rotation-derived age distributions of its stars (left) and planet
  hosts (middle).  The top row shows all stars with temperatures of
  3800--6200\,K for which we calculated rotational ages.  Opaque lines
  impose quality cuts on binarity, crowding, and the star's
  evolutionary state; transparent lines do not.  The bottom row shows
  stars with temperatures of 4400--5400\,K, which have more precise
  ages due to their fast spin-down.  The statistical uncertainty for a
  mean star at 1, 2, and 3\,Gyr is shown, and is identical across each
  row.  Finally, the right panel compares the rotation-derived ages
  against star formation histories derived using CMD fitting (dashed
  and dotted lines).  Completeness in Kepler's $P$$_{\rm rot}$
  detection sensitivity is near unity until $t$$\lesssim$3\,Gyr
  \citep{2022ApJ...937...94M}.
  \label{fig:hist_tgyro}
  }
\end{figure*}

@@

The similarity of the KOI host and Kepler star age distributions adds
some nuance to the argument that young transiting planets are hard to
detect due to the photometric variability of their host stars.  If
true, this statement seems to hold only for the very youngest
($\lesssim$0.4\,Gyr) stars, where there is a marginal deficit in the
KOI host age distribution relative to the parent stellar sample.
Examining scatter plots of the planet properties as a function of
time, we similarly find that fewer systems at $t_{\rm
gyro}$$<$0.4\,Gyr are detected with $P_{\rm orb}$$\gtrsim$30\,days
than at $t_{\rm gyro}$$>$0.4\,Gyr.

An important separate selection effect concerns Kepler's ability to
detect the rotation signals.  \citet{2022ApJ...937...94M} studied this
issue, and found for Sun-like stars that the fraction of rotation
signals detectable by Kepler is near unity up to $\approx$3\,Gyr, and
that it drops to almost zero by $\approx$5\,Gyr.  This trend is
opposite in functional form to our derived age distributions.

@@

Let us focus on blemish-free stars with blemish-free planets: $Q_{\rm
star}$ must not have any of bits zero through nine raised, and $Q_{\rm
planet}$ must similarly have no quality flags raised.  Under this
constraint, our results include \ltonegyrhighqconfirmedtwosided\
confirmed planets younger than 1\,Gyr with two-sided $t_{\rm gyro}$
and $t_{\rm Li}$, and \ltonegyrhighqconfirmedonesided\ confirmed
planets with comparable ages for which $t_{\rm gyro}$ is two-sided
while $t_{\rm Li}$ is one-sided.

@@

We were interested in not only the overall age distribution of the
Kepler field, but also in the reliability of individual ages for young
stars known to host planets.  We attempted to assess this reliability using
quality bitmasks (Table~\ref{tab:planets}), a comparison against \deleted{open }
cluster ages (Section~\ref{sec:litagecomparison}), and a comparison
between rotation and lithium ages.  The \deleted{open }cluster comparison
suggested that our ages were generally accurate at $\ll$0.5\,dex
across 0.04--2.5\,Gyr, comparable to their quoted precision.
Nonetheless, the upper-left panel of Figure~\ref{fig:agescalecompone}
does show outliers, typically $\lesssim$10\% of the population
in any cluster.  These stars could be either field star contaminants,
or else anomalous stars whose physical rotation histories were altered
by processes not captured by our statistical uncertainties.
Similarly, although we found $t_{\rm gyro}$ and $t_{\rm Li}$ to agree
for \ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
stars, we did find one system, Kepler-786, with radically different
lithium (\kepseveneightsix\,Myr) and rotational
(\kepseveneightsixgyro\,Myr) ages.  Planetary mergers
are one process that could produce such a signal, but
testing this would require a method for measuring differential abundances
across a large number of elements.

\added{A separate plausible origin for rotational outliers is that they come from
erroneously reported rotation periods.  While
the large surveys do show good internal consistency (Appendix~\ref{app:mcqonly}),
the referee correctly noted that this only a comparative test of methods,
rather than an ability to actually recover ground truth \citep[e.g.][]{Aigrain2015}.
One could imagine an adversarial scenario in which, say, 10-20\% of stars
with $P_{\rm rot}$$>$20\,days are erroneously reported, and in fact have much
longer rotation periods.  In this scenario, our analysis would overestimate the
rate by which the SFR in the Kepler field has decreased, by 7-14\%. }

@@

\startlongtable
\begin{deluxetable*}{lllllllrrrrrr}
  \tabletypesize{\scriptsize}
  \tablecaption{
  Ages of Kepler target stars derived from rotation periods.  The
  full machine-readable table includes
  \nuniqstarfinitegyroage\ stars with
  finite reported gyrochrone ages.  The quality
  bitmask is as in Table~\ref{tab:planets}; requiring bits zero to
  nine to be null yields \nuniqstarsantosrotgyroappl\ stars for
  which gyrochronology is likely to be valid.  \label{tab:stars}}
  %\toprule
  %\midrule
  %\endhead
  \startdata
  KIC & Gaia DR3 &  $T_{\rm eff}$ & $P_{\rm rot}$ & $t_{\rm gyro}$ & $Q_{\rm star}$ \\
  -- &   -- & K & days &  Myr &    int  \\
  \hline
  \input{table_star_gyro} %\\
  %\hline
  \enddata
% \tablecomments{This table includes X, Y, Z... The machine-readable version,
%   available online, includes additional columns for .... 
% }
\end{deluxetable*}

@@

This analysis is particularly interesting in relation to the question
of how fast stars $\gtrsim$4\,Gyr old spin down.  If we lived in a
universe in which stellar rotation rates and spot-induced photometric
amplitudes remained {\it constant} after $\approx$3\,Gyr, then older
stars would tend to contaminate the 2.5-3\,Gyr peaks in
Figures~\ref{fig:hist_tgyro} and~\ref{fig:hist_tgyro_mcq}.  This
scenario deserves consideration, given evidence from asteroseismic
rotation periods that has pointed to slowed spin-down rates after
$\approx$4\,Gyr for Sun-like stars
\citep[e.g.][]{vanSaders_2016,2024ApJ...962..138S}.  However, because
the \citeauthor{McQuillan_2014} sample does not include many stars
older than $\approx$4\,Gyr \citep{2022ApJ...937...94M}, this analysis
would argue that the 2.5-3\,Gyr peak cannot be caused by a pile-up of
much older stars with especially lethargic spin-down rates.

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.43\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
		}
		\subfloat{
			\includegraphics[width=0.43\textwidth]{prot_teff_McQuillan2014only_dquality.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Variant of Figure~\ref{fig:prot_vs_teff}}, comparing the default
    rotation periods ({\it left}) from \citet{Santos_2019,Santos_2021}
    against the \citet{McQuillan_2014} rotation periods ({\it right}).
    All stars are required to be amenable to rotation-based age dating
    (i.e.~none of bit flags zero through nine raised).
	}
	\label{fig:prot_vs_teff_mcq}
\end{figure*}

>>>>

Finally, since our scope is focused on rotation-based ages, we
restricted our attention to stars with reported $P_{\rm
rot}$$<$45\,days.  The slowest-rotating FGK stars in the open clusters
used to calibrate our gyrochronology model have $P_{\rm
rot}$$\approx$35\,days.  Figure~\ref{fig:prot_vs_teff} shows a subset
of the resulting \nuniqstarsantosrot\ Kepler stars with rotation
periods compiled from \citetalias{Santos_2019},
\citetalias{Santos_2021}, and our extended KOI list.\added{  The gray
lines in this figure are the polynomials described by \citet{Bouma_2023}
that 
fit the slow rotation sequences of the Pleiades \citep{Rebull_2016},
Praesepe \citep{Rampalli_2021}, NGC-6811 \citep{Curtis_2019_ngc6811}, NGC-6819 \citep{Meibom_2015},
Ruprecht-147 \citep{Curtis_2020}, and M67 \citep{Barnes2016,Dungee_2022,Gruner_2023}.}

@@

\subsection{Stellar Quality Flags}
\label{subsec:flags}
We calculated gyrochrone ages for all \nuniqstarsantosrotteffcut\
stars with reported rotation periods that had effective temperatures
of 3800--6200\,K.  To identify stars for which we suspect these ages
may not be valid, we then built a set of quality flags which we
condensed into a single binary number: $Q_{\rm star}$.  When and how
this bitmask should be applied depends on the question being asked.
If the goal is to construct a false-positive free sample, all the
quality flags could be applied.  If the goal is to construct a
complete sample, then consider the examples of Kepler~1627Ab
($\approx$40\,Myr) and Kepler~51c ($\approx$625\,Myr).  The former has
a high RUWE due to a resolved binary companion \citep{Bouma_2022a};
the latter is on a grazing orbit \citep{2014ApJ...783...53M}.  We
leave selection for or against such cases as a decision to the user.
For our own analysis, we assume that a star is suitable for
gyrochronology if none of bits zero through \replaced{nine}{ten} (inclusive) are
raised.  For analyses that require all stellar rotation periods to
come from the same detection pipeline, we further require bit \replaced{11}{12} to be
zero.

@@

%The rotation period
%distribution of sources excluded in this manner includes a relatively
%larger fraction of ``rapid rotators'' than the main sequence sources,
%likely due to a combination of astrophysical and systematic effects.
%See for instance Figure~8a of \citet{2022AJ....164..137K} or
%Figures~12 and 13 of \citet{2023ApJS..268....4F} for examples of how
%the period distribution of this population differs from stars near the
%main sequence.
\added{{\it Metallicity (bit 10)}---Theoretical models
	suggest that stars with non-solar metallicities
	spin down at different rates than solar-metallicity stars 
	\citep{Amard2020}.   This expectation has been hard
	to verify due to a dearth of nearby non-solar metallicity open clusters.
	However, if true, this effect could cause systematic
	biases for gyrochrone ages \citep[e.g.~Figure~9 of][]{Claytor2020}.
	We therefore flagged stars with spectroscopic $|[{\rm Fe/H}]|$$>$0.3
	from either LAMOST DR5 \citep{Zong2018} or CKS \citep{Petigura_2022}
	as being outside the range of our gyrochronology calibration.
	We discuss limitations of this approach in Section~\ref{sec:disc}.
}

{\it Candidate pulsators and close-in binaries (bit
\replaced{10}{11})}---\citeauthor{Santos_2021} included in a flag for candidate
``classical pulsators'' (e.g.\ Cepheids) and close-in binaries.
Visual inspection shows that although this flag selects many bona fide
objects in these classes, it also selects most known young planets
with $P_{\rm rot}$$\lesssim$5\,day.  These objects are
neither classical pulsators nor close binaries.  While we propagated
this flag into our set of quality flags, we therefore generally ignore
it.

{\it Not in the homogeneous stellar rotation sample (bit \replaced{11}{12})}--- This
flag was set for the KOIs with rotation periods drawn from a source
other than the \citeauthor{Santos_2019} pipeline.

@@

	\vspace{-0.35cm}
    \subfloat{
        \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000_preciseagesonly.pdf}
    }
  \end{center}
  \vspace{-0.6cm}
  \caption{{\bf Kepler's demographic cliff,} visible in the
  rotation-derived age distributions of its stars (left) and planet
  hosts (middle).  The top row shows all stars with temperatures of
  3800--6200\,K for which we calculated rotational ages.  Opaque lines
  impose quality cuts on binarity, \added{metallicity, }crowding, and the star's
  evolutionary state; transparent lines do not.  The bottom row shows
  stars with temperatures of 4400--5400\,K, which \replaced{have more precise
  ages due to their fast spin-down}{often have more precise ages}.  The statistical uncertainty for a
  mean star at 1, 2, and 3\,Gyr is shown, and is identical across each
  row.  Finally, the right panel compares the rotation-derived ages
  against star formation histories derived using CMD fitting (dashed
  and dotted lines).  Completeness in Kepler's $P$$_{\rm rot}$
  detection sensitivity is near unity until $t$$\lesssim$3\,Gyr
  \citep{2022ApJ...937...94M}.
  \label{fig:hist_tgyro}
  }
\end{figure*}

@@

The similarity of the KOI host and Kepler star age distributions adds
some nuance to the argument that young transiting planets are hard to
detect due to the photometric variability of their host stars.  If
true, this statement seems to hold only for the very youngest
($\lesssim$0.4\,Gyr) stars, where there is a marginal deficit in the
KOI host age distribution relative to the parent stellar sample.
Examining scatter plots of the planet properties as a function of
age, we similarly find that fewer systems at $t_{\rm
gyro}$$<$0.4\,Gyr are detected with $P_{\rm orb}$$\gtrsim$30\,days
than at $t_{\rm gyro}$$>$0.4\,Gyr.

An important separate selection effect concerns Kepler's ability to
detect the rotation signals.  \citet{2022ApJ...937...94M} studied this
issue, and found for Sun-like stars that the fraction of rotation
signals detectable by Kepler is near unity up to $\approx$3\,Gyr, and
that it drops to almost zero by $\approx$5\,Gyr.  This trend is
opposite in functional form to our derived age distributions\replaced{.}{and
	most of the stars in our sample are indeed G dwarfs (see e.g.,~the $t_{\rm gyro}$--$T_{\rm eff}$ distributions 
	in Appendix~\ref{app:age_diagnostic}).}

@@

Let us focus on blemish-free stars with blemish-free planets: $Q_{\rm
star}$ must not have any of bits zero through \replaced{nine}{ten} raised, and $Q_{\rm
planet}$ must similarly have no quality flags raised.  Under this
constraint, our results include \ltonegyrhighqconfirmedtwosided\
confirmed planets younger than 1\,Gyr with two-sided $t_{\rm gyro}$
and $t_{\rm Li}$, and \ltonegyrhighqconfirmedonesided\ confirmed
planets with comparable ages for which $t_{\rm gyro}$ is two-sided
while $t_{\rm Li}$ is one-sided.

@@

We were interested in not only the overall age distribution of the
Kepler field, but also in the reliability of individual ages for young
stars known to host planets.  We attempted to assess this reliability using
quality bitmasks (Table~\ref{tab:planets}), a comparison against \deleted{open }
cluster ages (Section~\ref{sec:litagecomparison}), and a comparison
between rotation and lithium ages.  The \deleted{open }cluster comparison
suggested that our ages were generally accurate at $\ll$0.5\,dex
across 0.04--2.5\,Gyr, similar to their quoted precision.
Nonetheless, the upper-left panel of Figure~\ref{fig:agescalecompone}
does show outliers, typically $\lesssim$10\% of the population
in any cluster.  These stars could be either field star contaminants,
or else anomalous stars whose physical rotation histories were altered
by processes not captured by our statistical uncertainties.
Similarly, although we found $t_{\rm gyro}$ and $t_{\rm Li}$ to agree
for \ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
stars, we did find one system, Kepler-786, with radically different
lithium (\kepseveneightsix\,Myr) and rotational
(\kepseveneightsixgyro\,Myr) ages.  Planetary mergers
are one process that could produce such a signal, but
testing this would require a method for measuring differential abundances
across a large number of elements.

\added{A separate plausible origin for rotational outliers is that they come from
erroneously reported rotation periods.  While
the large surveys do show good internal consistency (Appendix~\ref{app:mcqonly}),
the referee correctly pointed out that this only a comparative test of methods,
rather than an ability to recover ground truth \citep[e.g.][]{Aigrain2015}.
One could imagine an adversarial scenario in which, say, 10-20\%\ of stars
with $P_{\rm rot}$$>$20\,days are erroneously reported, and in fact have much
longer rotation periods.  In this scenario, our analysis would overestimate the
rate by which the SFR in the Kepler field has decreased, by 7-14\%. }

@@

\startlongtable
\begin{deluxetable*}{lllllllrrrrrr}
  \tabletypesize{\scriptsize}
  \tablecaption{
  Ages of Kepler target stars derived from rotation periods.  The
  full machine-readable table includes
  \nuniqstarfinitegyroage\ stars with
  finite reported gyrochrone ages.  The quality
  bitmask is as in Table~\ref{tab:planets}; requiring bits zero to
  ten to be null yields \nuniqstarsantosrotgyroappl\ stars for
  which gyrochronology is likely to be valid.  \label{tab:stars}}
  %\toprule
  %\midrule
  %\endhead
  \startdata
  KIC & Gaia DR3 &  $T_{\rm eff}$ & $P_{\rm rot}$ & $t_{\rm gyro}$ & $Q_{\rm star}$ \\
  -- &   -- & K & days &  Myr &    int  \\
  \hline
  \input{table_star_gyro} %\\
  %\hline
  \enddata
% \tablecomments{This table includes X, Y, Z... The machine-readable version,
%   available online, includes additional columns for .... 
% }
\end{deluxetable*}

@@

This analysis is particularly interesting in relation to the question
of how fast stars $\gtrsim$4\,Gyr old spin down.  If we lived in a
universe in which stellar rotation rates and spot-induced photometric
amplitudes remained {\it constant} after $\approx$3\,Gyr, then older
stars \replaced{would tend to }{could }contaminate the 2.5-3\,Gyr peaks in
Figures~\ref{fig:hist_tgyro} and~\ref{fig:hist_tgyro_mcq}.  This
scenario deserves consideration, given\replaced{ evidence from asteroseismic
rotation periods that has pointed to}{ {\it i)} asteroseismic evidence for} slowed spin-down rates after
$\approx$4\,Gyr for Sun-like stars
\citep[e.g.][]{vanSaders_2016,2024ApJ...962..138S}\added{, and {\it ii)} the observation of a ``long-period edge''
in the $P_{\rm rot}$-$T_{\rm eff}$ distribution which
may pile up over similar timescales \citep{David_2022}}.  However, \replaced{because
the \citeauthor{McQuillan_2014} sample does not include many stars
older than $\approx$4\,Gyr \citep{2022ApJ...937...94M}, this analysis}{the analysis
by \citet{2022ApJ...937...94M} fitted isochrones to a high-quality subset of Kepler data,
and found that the
sample of detected rotators from \citeauthor{McQuillan_2014} does not include many
stars older than $\approx$\,4\,Gyr.}
\replaced{would argue}{This suggests} that the 2.5-3\,Gyr peak \replaced{cannot}{is unlikely to} be caused by a pile-up of
much older stars with especially lethargic spin-down rates.

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.43\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
		}
		\subfloat{
			\includegraphics[width=0.43\textwidth]{prot_teff_McQuillan2014only_dquality.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Variant of Figure~\ref{fig:prot_vs_teff}}, comparing the default
    rotation periods ({\it left}) from \citet{Santos_2019,Santos_2021}
    against the \citet{McQuillan_2014} rotation periods ({\it right}).
    All stars are required to be amenable to rotation-based age dating
    (i.e.~none of bit flags zero through \replaced{nine}{ten} raised).
	}
	\label{fig:prot_vs_teff_mcq}
\end{figure*}
